A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2020-05-19 manuscript.Rmd
further revisions

Here, we combine four software tools, whose interplay can guarantee full computational reproducibility of data analyses and their reporting. There are various ideas on how to enhance reproducibility [@piccoloToolsTechniquesComputational2016], four of which we believe to be particularly important: dynamic document generation: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management [@askrenUsingMakeReproducible2016], and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. We argue that only a workflow using all four concepts in unison can guarantee confidence in reproducing a scientific report [see @turingway2020 for similar arguments]. Various implementations of these concepts exist, but we consider the following four best suited for analyses centered on the R environment [@r2019] but also allowing for external dependencies: R Markdown [@rmarkdown2019] for dynamic document generation, Git [@chaconProGit2014] for version control, Make [@feldmanMakeProgramMaintaining1979] for dependency management, and Docker [@merkelDockerLightweightLinux2014] for containerization. Each of these software solutions serves a valuable meta-scientific goal (reproducibility) and increases the researches' productivity. They all are highly flexible and powerful, so that mastering them fully, requires a considerable amount of practice. However, for our purposes, it is sufficient to master a valuable minimal subset of functions to ensure the reproducibility of scientific analyses.

@@

The first line after the comment is the first (default) target called 'all', which depends on manuscript.pdf, which itself is a target. If Make is called without an argument, the first target is built. To create `manuscript.pdf` (the second target in the file), the file `manuscript.Rmd` needs to be rendered, which depends on `data/ìris_prepped.csv`. This dependency is itself a target (the third target in the file). To create `data/ìris_prepped.csv`, `R/prepare_data.R` and `data/iris.csv` are needed. If you type `make manuscript.pdf`, Make first checks whether the dependencies do exist and, if not, creates them. Here, if `data/ìris_prepped.csv` does not exist, Make creates it by executing the third target. Also, if one of the dependencies of a target is newer than the target itself, Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer than `data\ìris_prepped.csv`, Make will attempt to recreate `data\ìris_prepped.csv` first and then `manuscript.pdf`. If there is a dependency missing, and there is no target to make it, Make stops with an error message. It is a convention to have the first target named `all`, which depends on everything. Subsequently, the command `make` without any argument automatically creates everything possible in the project. The button `Build All` from within RStudio triggers this process (see Figure \ref{fig:buildpane-rstudio}).

```{r buildpane-rstudio, fig.cap="Build pane in RStudio with access to Makefile target 'all'."}
knitr::include_graphics("Images/build-pane.png")
```

@@

Docker is a tool that allows encapsulation, sharing, and re-creation of a computational environment on most operating systems (Windows, macOS & Linux). Docker achieves these goals by setting up a virtual computer, on which it can execute a well-defined series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands on the virtual computer, e.g., running `Rscript` or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has Docker installed. Irrespective of the machine that is executing the container, the computational environment is the same for the programs running inside the container. The most important advantage over traditional virtual machines is that containers are lightweight; that is, they start rapidly and do not need much storage space. Docker achieves this by reusing large parts of the host's operating system (the extend varies between Linux, macOS and Windows).

With the following example, we demonstrate the importance of storing computational environment. Generally, with containers, we would like to safeguard against changes to the computational environment resulting in unexpected consequences, e.g., changes in the functionality or default options in packages or even in the R environment itself. While the R programming language is considered stable and much effort is put into backward compatibility, even basic functions like `sample()` (to randomly sample from a set) sometimes change their behaviour from one version to another. To ensure reproducibility in analyses based on a computer's pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly. Consider the following R code to randomly draw five numbers between 1 and 10:

@@

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R Version (due to `set.seed()`). Using Docker, we can start an image which contains the R Version 3.5.0, and execute the code there.

@@

Only rarely does an analysis depend on base R only. Typically, a considerable number of packages is required that each may depend on multiple other packages, and the addition of each package will increase the likelihood of breaking reproducibility with every update to any of the packages or base R (the resulting frustration is sometimes referred to as _dependency hell_). The whole endeavour of reproducibility is therefore at stake every time an update is rolled out. To ensure long-term reproducibility, Docker replicates the original computational environment of an analysis exactly. Note, that we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; our point is that full computational reproducibility is only achieved if the software versions used originally are precisely documented. Among other things, this makes it possible to trace back update histories to discover which change in which package caused the non-reproducibility. Quite to the contrary, with containerization, it gets easier than ever to safely update to new versions just by changing the R version number of the Docker image (and reverting back if this update breaks code). This convenience is possible because of the efforts of the [Rocker project](https://GitHub.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a], which provides Docker images pre-configured with an installation of selected R versions. These packages are taken from MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R packages fixed to the last date on which the R version of the image was the most recent. Building upon these Rocker images, researchers can easily build their own Docker images with all required R packages. The rocker project also provides images that include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the R Markdown package with LaTeX (`rocker/verse`). Because our workflow relies on R Markdown, we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [Dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

# execute a given command in a given container image; do not save
# the state of the container; accept inputs from and return outputs to terminal
docker run --rm -it <IMAGENAME> <COMMAND>
```

@@

# start an interactive R session in the container named 'reproducible-research'
docker run --rm -it reproducible-research R
```

@@

project := $(notdir $(CURDIR))
current_dir := $(CURDIR)
home_dir := $(current_dir)
uid = --user $(shell id -u)

ifeq ($(DOCKER),TRUE)
    run:=docker run --rm --user $(uid) -v $(home_dir):/home/rstudio $(project)
    current_dir=/home/rstudio
endif

all: manuscript.pdf

build: Dockerfile
    docker build -t $(projekt) .

manuscript.pdf: manuscript.Rmd reproducible-research.bib
    $(run) Rscript -e 'rmarkdown::render("$(current_dir)/$<")'
```

@@

Finally, we conclude with some notes on project organization, which we think makes migrating projects to a reproducible workflow easier. The first step towards reproducibility is to create an R script or R Markdown file as the primary entry point for the analysis that runs on a local computer without error and performs the main statistical analyses. Next, one needs to make sure that all files relevant to the analysis can be moved to another computer. To this end, it is recommendable that all files are within one folder (or enclosed subfolders within it) and all paths are relative to that folder because absolute paths are specific to a given computer. A robust solution to the problem of making sure that file access does not break across computing platforms are [RStudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` package [@here2017] to manage file access. The here package solves two common issues with relative paths. First, it takes care of the fact that path separator characters vary across operating systems (typically, slash or backslash). Second, anchor points of relative paths may differ depending on context. For example, knitr interprets paths relative to the dynamic document, whereas R has a current working directory that may change over the course of an R session. The `here` package provides consistent paths relative to the project directory. The following three examples refer to local files ranging from absolute paths with system-specific path separators (bad) to relative paths using the `here` package:

@@

The folder where all the files reside that you need for analysis (code and data), is referred to as a 'project' (or sometimes as a 'research compendium'). Working with projects is particularly convenient with RStudio, an integrated development environment (IDE) for R. It is useful to organize a data analysis project in a way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 in @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

@@

The overarching goal of this paper was to provide a complete and easy-to-use workflow that allows confidence in the reproducibility of R-based data analyses. Analyses following our workflow can be reproduced with four commands (here shown for our paper):

@@

That is, this reproduces a scientific report exactly without regard to the user, the operating system, the software, the timepoint, or interim changes to the involved files. To that end, the proposed workflow relies on tools that have been the foundation of reliable software development for years or even decades. As a by-product, it makes transparent how statistical results depend on the software that created them and, by virtue of this transparency, facilitates later reuse by other researchers.

Each tool in the workflow reduces the chances of non-reproducibility. Dynamic reporting with R Markdown guarantees consistency between computational results and their reporting; Version control with Git ensures permanence and consistency across multiple versions of data and code; Dependency management with Make affirms defined entry-points and dependency resolution; containerization with Docker secures reproducibility of the full computational environment. We believe that this tightening of loose ends does not restrict researchers but enables them to operate on a solid basis to deliver sound and sustainable research.

@@

While our approach was designed to scale well with the complexity of a computationally intense project, we realize that this flexibility may not be straightforward to integrate into a researcher's everyday workflow. There are R packages that implement parts of our workflow and, thus, lower the threshold for adoption when the full flexibility provided by our workflow is not needed. The use of R Markdown within a project, tracked with Git can be simplified with the [`workflowR`](https://github.com/jdblischak/workflowr) package [@workflowr]. The [`drake`](https://github.com/ropensci/drake) package [@drake] is directly inspired by Make and takes an R centric approach, making it especially suited for projects only involving R, but can also handle external dependencies. The [liftR ](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr] and the [`holepunch`](https://karthik.github.io/holepunch/) package [@holepunch] automatize the use of Docker. The former is perfectly compatible with the described workflow, and we recommend it to users who are not comfortable with command-line use of Docker. `holepunch` uses [binder](https://mybinder.org) [@binder2018] to move the analysis to the cloud, so that no local installation of Docker is required. `holepunch` is well suited for simple analyses with low computational demands because binder's memory and computing time is limited. There are several alternatives to Docker that manage dependencies on R packages. [`renv`](https://rstudio.github.io/renv/articles/renv.html) [@renv] is a way to freeze package version via local copies of packages in the project, but it does not guarantee a given base R version or system dependencies. Similar approaches are taken by [`jetpack`](https://github.com/ankane/jetpack) [@jetpack], [`miniCRAN`](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [`checkpoint`](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. The package [`reprex`](https://github.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex] is also worth noting, but its scope is limited. A particularly noteworthy approach is provided by the [`worcs`](https://cran.r-project.org/web/packages/worcs) package [@worcs, @worcspaper], which is an R project template that creates a standardized file structure for code and data supporting version management with Git, package management with renv and dynamic document generation with R Markdown. We acknowledge that worcs is much easier to install and provides a one-click solution for the creation of reproducible projects. It achieves a high standard of reproducibility but does not guarantee full computational reproducibility and is limited to dependency management within the R environment.

>>>>

Here, we combine four software tools, whose interplay can guarantee full computational reproducibility of data analyses and their reporting. There are various ideas on how to enhance reproducibility [@piccoloToolsTechniquesComputational2016], four of which we believe to be particularly important: dynamic document generation: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management [@askrenUsingMakeReproducible2016], and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. We argue that only a workflow using all four concepts in unison can guarantee confidence in reproducing a scientific report [see @turingway2020 for similar arguments]. Various implementations of these concepts exist, but we consider the following four best suited for analyses centered on the R environment [@r2019] but also allowing for external dependencies: R Markdown [@rmarkdown2019] for dynamic document generation, Git [@chaconProGit2014] for version control, Make [@feldmanMakeProgramMaintaining1979] for dependency management, and Docker [@merkelDockerLightweightLinux2014] for containerization. Each of these software solutions serves a valuable meta-scientific goal (reproducibility) and increases the researches' productivity. They all are highly flexible and powerful, so that mastering them fully, requires a considerable amount of practice. However, for our purposes, it is sufficient to master a valuable minimal subset of functions to ensure the reproducibility of scientific analyses. We recommend using RStudio, an integrated development environment (IDE) for R, which provides simplified access to essential features of some of the tools.

@@

The first line after the comment is the first (default) target called 'all', which depends on manuscript.pdf, which itself is a target. If Make is called without an argument, the first target is built. To create `manuscript.pdf` (the second target in the file), the file `manuscript.Rmd` needs to be rendered, which depends on `data/ìris_prepped.csv`. This dependency is itself a target (the third target in the file). To create `data/ìris_prepped.csv`, `R/prepare_data.R` and `data/iris.csv` are needed. If you type `make manuscript.pdf`, Make first checks whether the dependencies do exist and, if not, creates them. Here, if `data/ìris_prepped.csv` does not exist, Make creates it by executing the third target. Also, if one of the dependencies of a target is newer than the target itself, Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer than `data\ìris_prepped.csv`, Make will attempt to recreate `data\ìris_prepped.csv` first and then `manuscript.pdf`. If there is a dependency missing, and there is no target to make it, Make stops with an error message. It is a convention to have the first target named `all`, which creates the entire project. Subsequently, the command `make` without any argument automatically creates everything possible in the project. The button `Build All` from within RStudio triggers this process (see Figure \ref{fig:buildpane-rstudio}).

```{r buildpane-rstudio, eval=TRUE, echo=FALSE, fig.cap="Build pane in RStudio with access to Makefile target 'all'."}
knitr::include_graphics("Images/build-pane.png")
```

@@

Docker is a tool that allows encapsulation, sharing, and re-creation of a computational environment on most operating systems (Windows, macOS, & Linux). Docker achieves these goals by setting up a virtual computer, on which it can execute commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands on the virtual computer, e.g., running `Rscript` or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has Docker installed. Regardless of the machine that is executing the container, the computational environment is identical for the programs running inside the container. The most important advantage over traditional virtual machines is that containers are lightweight: they start rapidly, run with little overhead, and do not need much storage space. Docker achieves this by reusing large parts of the host's operating system (the extent varies between Linux, macOS and Windows).

With the following example, we demonstrate the importance of documenting and storing the computational environment. Generally, with containers, we would like to safeguard against changes to the computational environment resulting in unexpected consequences, e.g., changes in the functionality or default options in packages or even in the R environment itself. While the R programming language is considered stable and much effort is put into backward compatibility, even basic functions like `read.csv()` (to load data) or `sample()` (to randomly sample from a set) sometimes change their behaviour from one version to another. For example, to ensure reproducibility of analyses based on a computer's pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly. Consider the following R code to randomly draw five numbers between 1 and 10:

@@

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R Version (because of `set.seed()`). Using Docker, we can start an image which contains the R Version 3.5.0, and execute the code there.

@@

Only rarely does an analysis depend on base R only. Typically, a considerable number of packages is required that each may depend on multiple other packages. Each update of each package in this dependency hierarchy and updates to base R itself will increase the likelihood of breaking reproducibility (the resulting frustration is sometimes referred to as _dependency hell_). The whole endeavour of reproducibility is therefore at stake every time an update is rolled out. To ensure long-term reproducibility, our worflow replicates the original computational environment of an analysis exactly. Note, that we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; our point is that full computational reproducibility is only achieved if the software versions used originally are precisely documented. Among other things, this makes it possible to trace back update histories to discover which change in which package caused the non-reproducibility. Quite to the contrary, with containerization, it gets easier than ever to safely update to new versions just by changing the R version number of the Docker image (and reverting back if this update breaks code). This convenience is possible because of the efforts of the [Rocker project](https://GitHub.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a], which provides Docker images pre-configured with an installation of selected R versions. These packages are taken from MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R packages fixed to the last date on which the R version of the image was the most recent. Building upon these Rocker images, researchers can easily build their own Docker images with all required R packages. The rocker project also provides images that include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the R Markdown package with LaTeX (`rocker/verse`). Because our workflow relies on R Markdown, we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [Dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

# execute a command in a container image; do not save the state 
# of the container; accept inputs from and return outputs to terminal
docker run --rm -it <IMAGENAME> <COMMAND>
```

@@

# start an interactive R session in the 
# container named 'reproducible-research'
docker run --rm -it reproducible-research R
```

@@

# set local variables for later use
project := $(notdir $(CURDIR))
current_dir := $(CURDIR)
home_dir := $(current_dir)
uid = --user $(shell id -u)

# determine if DOCKER=TRUE was given
# if so, run everything in docker
# if not, run everything locally
ifeq ($(DOCKER),TRUE)
    run:=docker run --rm --user $(uid) -v $(home_dir):/home/rstudio $(project)
    current_dir=/home/rstudio
endif

# default target is target 'manuscript.pdf'
all: manuscript.pdf

# build the docker container
build: Dockerfile
    docker build -t $(project) .

# build manuscript.pdf from Rmd file
# run in docker if DOCKER=TRUE else locally
manuscript.pdf: manuscript.Rmd reproducible-research.bib
    $(run) Rscript -e 'rmarkdown::render("$(current_dir)/$<")'
```

@@

Finally, we conclude with some notes on project organization, which we think makes migrating projects to a reproducible workflow easier. The first step towards reproducibility is to create an R script or R Markdown file as the primary entry point for the analysis that runs on a local computer without error and performs the main statistical analyses. Next, one needs to make sure that all files relevant to the analysis can be moved to another computer. To this end, it is recommended that all files reside within a one folder (or enclosed subfolders within it) and all paths are relative to that folder because absolute paths are specific to a given computer. A robust solution to the problem of making sure that file access does not break across computing platforms are [RStudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` package [@here2017] to manage file access. The here package solves two common issues with relative paths. First, it takes care of the fact that path separator characters vary across operating systems (typically, slash or backslash). Second, it solves the issue that anchor points of relative paths may differ depending on context. For example, knitr interprets paths relative to the dynamic document, whereas R has a current working directory that may change over the course of an R session. The `here` package provides consistent paths relative to the project directory. The following three examples refer to local files ranging from absolute paths with system-specific path separators (bad) to relative paths using the `here` package:

@@

The folder where all the files reside that you need for analysis (code and data), is referred to as a 'project' (or sometimes as a 'research compendium'). Working with projects is particularly convenient with RStudio. It is useful to organize a data analysis project in a way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 in @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

@@

The overarching goal of this paper was to provide a complete and easy-to-use workflow that allows confidence in the reproducibility of R-based data analyses. Analyses following our workflow can be reproduced with four commands (here shown for this manuscript):

@@

The workflow enables reproduction a scientific report exactly without regard to the local operating system, locally installed software, time, or interim changes to the project files. To that end, the proposed workflow relies on tools that have been the foundation of reliable software development for years or even decades. As a by-product, it makes transparent how statistical results depend on the software that created them and, by virtue of this transparency, facilitates later reuse by other researchers.

Each tool in the workflow reduces the chances of non-reproducibility. Dynamic reporting with R Markdown guarantees consistency between computational results and their reporting; Version control with Git ensures permanence and consistency across multiple versions of data and code; Dependency management with Make provides defined entry-points while mapping out dependencies between all components of a project; containerization with Docker guarantuees full computational reproducibility. We believe that the proposed combination of tools does not restrict researchers but enables them to operate on a solid basis to deliver transparent and sustainable research.

@@

While our approach was designed to scale well with the complexity of a computationally intense project, we realize that this flexibility may not be straightforward to integrate into researchers' everyday workflow. There are various R packages that implement parts of our workflow and, thus, lower the threshold for adoption when the full flexibility provided by our workflow is not needed. The use of R Markdown within a project, tracked with Git can be simplified with the [`workflowR`](https://github.com/jdblischak/workflowr) package [@workflowr]. The [`drake`](https://github.com/ropensci/drake) package [@drake] is directly inspired by Make and takes an R centric approach, making it especially suited for projects only involving R, but can also handle external dependencies. The [liftR ](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr] and the [`holepunch`](https://karthik.github.io/holepunch/) package [@holepunch] automatize the use of Docker. The former is perfectly compatible with the described workflow, and we recommend it to users who are not comfortable with command-line use of Docker. `holepunch` uses [binder](https://mybinder.org) [@binder2018] to move the analysis to the cloud, so that no local installation of Docker is required. `holepunch` is well suited for simple analyses with low computational demands because binder's memory and computing time is limited. There are several alternatives to Docker that manage dependencies on R packages. [`renv`](https://rstudio.github.io/renv/articles/renv.html) [@renv] is a way to freeze package version via local copies of packages in the project, but it does not guarantee a given base R version or system dependencies. Similar approaches are taken by [`jetpack`](https://github.com/ankane/jetpack) [@jetpack], [`miniCRAN`](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [`checkpoint`](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. The package [`reprex`](https://github.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex] is also worth noting, but its scope is limited. A particularly noteworthy approach is provided by the [`worcs`](https://cran.r-project.org/web/packages/worcs) package [@worcs; @worcspaper], which is an R project template that creates a standardized file structure for code and data supporting version management with Git, package management with renv and dynamic document generation with R Markdown. We acknowledge that worcs is much easier to install and provides a one-click solution for the creation of reproducible projects. It achieves a high standard of reproducibility but does not guarantee full computational reproducibility and is limited to dependency management within the R environment.
